\thetitle

### 2023-06-03 lang-design.tex
Fix Grammar.

As analyzed earlier, ordinary recursors, such as "tm_rect", are impossible
within a family in which the name of the inductive type is late bound.
However,
a key observation is that extensible inductive types still admit a
weakened elimination principle where the motive is an \lsti{option} type.
For example, within family "STLC", the partial recursor for
"tm" has the following type, with extra equations describing
its behvaiour on each constuctors:

\begin{centered}
\begin{minipage}{\textwidth}
\newcommand\hlcase[1]{\tikzmarkin[disable rounded corners=true,set fill color=yellow!18,set border color=yellow!18]{#1}(0.00,-0.05)(-0.00,0.20)}
\begin{lstlisting}[basicstyle=\fontsize{8.25}{9.9}\ttfamily]
tm_prect_STLC : \forall (P : tm -> Type), [@\hlcase{tm-partialrec-case-unit}@]option (P tm_unit)[@\tikzmarkend{tm-partialrec-case-unit}@] -> [@\hlcase{tm-partialrec-case-var}@](\forall x, option (P (tm_var x)))[@\tikzmarkend{tm-partialrec-case-var}@] ->
  [@\hlcase{tm-partialrec-case-abs}@](\forall x t, option (P t) -> option (P (tm_abs x t)))[@\tikzmarkend{tm-partialrec-case-abs}@] ->
  [@\hlcase{tm-partialrec-case-app}@](\forall t1, option (P t1) -> \forall t2, option (P t2) -> option (P (tm_app t1 t2)))[@\tikzmarkend{tm-partialrec-case-app}@] -> \forall t, option (P t)
(* For arbitrary [@$P$@] [@$r_1$@] [@$r_2$@] [@$r_3$@] [@$r_4$@], and we denote [@"tmPR"@] := [@"tm_prect_STLC"@] [@$P$@] [@$r_1$@] [@$r_2$@] [@$r_3$@] [@$r_4$@] *)
(* We will have following equations *)
tm_unit_eq_STLC : tmPR tm_unit      = [@$r_1$@]
tm_var_eq_STLC  : tmPR (tm_var x)   = [@$r_2$@] x
tm_abs_eq_STLC  : tmPR (tm_abs x b) = [@$r_3$@] a b (tmPR b)
tm_app_eq_STLC  : tmPR (tm_app a b) = [@$r_4$@] a (tmPR a) b (tmPR b)
\end{lstlisting}
\end{minipage}
\end{centered}

"tm_prect_STLC" cannot be a trivial function even though its return type has "option". Because these four equations uniquely describe the "tm_prect_STLC" on the four constructors and make "tm_prect_STLC" behave similar to the ordinary dependent eliminator of the non-extensible version of "tm".

@@

This partial recursor seems non-canonical compared to the (dependent) eliminator for Inductive type as the latter can be considered as universal property. This extensionality makes dependent eliminator one of \textit{the most powerful} reasoning tools around inductive type.

Even though we cannot prove partial recursors are \textit{the
most powerful} reasoning tool for extensible inductive types,
the fact that propositional partial
recursors can derive injectivity already demonstrate itself as a good \textit{extensional characterization} of
(non-indexed) extensible inductive type (under certain
constraints)---because we can ``embed'' vanilla inductive types into
types supporting a partial recursor: a bit more formally, and
restricting our focus to non-indexed inductive types:

@@

  \begin{itemize}
    \item \textlabel{(Detectable Partiality)}{prop:detectable-partiality} for each pair $x : A_i \vdash B_i(x)$,
    \begin{minted}{Coq}
      Axiom detectable:
      forall {T} {a} (f: Bᵢ a → option T),
        {forall x, f x <> None} + {exists x, f x = None}.
    \end{minted} 
    \item Define $"C"$ as the inductive type using this list, i.e.,
    \begin{minted}{Coq}
Inductive C : Set := ... | cᵢ : forall (x : Aᵢ), (Bᵢ x → C) → C | ...
    \end{minted}
    \item Assume an arbitrary type \mintinline{Coq}{D : Set} with
    \begin{itemize}
      \item $n$ \textbf{functions} \mintinline{Coq}{dᵢ : forall (x : Aᵢ), (Bᵢ x → D) → D}  
      \item a partial recursor
      \begin{minted}[escapeinside=@@]{Coq}
prec : forall (R : Set), ..,
  (rᵢ : (forall (x : Aᵢ), (Bᵢ x → option R) → option R)), ..,@\YZ{Can R be indexed?}\EDJreply{This part actually currently incorrect and needs an extra constraint. Let me fix other parts first}\YZreply{Is this constraint Detectable Partiality?}@
      D → option R
      \end{minted}
      \item and $n$ (propositional) computational axioms: for all $i$, 
      \begin{minted}{Coq}
prec {R} r₁ r₂ .. rₙ (dᵢ aᵢ bᵢ) 
    = rᵢ aᵢ (fun x ⇒ prec {R} r₁ r₂ .. rₙ (bᵢ x))
      \end{minted}
    % where \mintinline{Coq}{(lift rᵢ) : (forall (x : Aᵢ), (Bᵢ x -> option R) -> option R)} is defined as expected
    \end{itemize}
  \end{itemize}
  Then there exists an \textbf{embedding} from "C" to "D".
  More concretely, we can have \mintinline{Coq}{inj : C → D} and 
  \mintinline{Coq}{linv : D → option C} defined using the eliminator of\/
  "C" and "prec" of\/ "D" (both acting like identity) such that
  \begin{minted}[escapeinside=@@]{Coq}
    forall c : C, linv (inj c) = some c@\YZ{Does 'linv' correpsond to 'reflect' above? If so, then the reader should be explicitly reminded of this fact rather than having to guess it.}\EDJreply{Added below. I mentioned left inverse 'linv' is the generalization}@
  \end{minted}
\end{theorem}
\begin{proof} 
First, for a fixed \mintinline{Coq}{x : Aᵢ}, given \mintinline{Coq}{fᵢ :  Bᵢ x → T},
\\ we have \mintinline{Coq}{lift fᵢ : Bᵢ x → option T}, simply wrapping option around the returning result.

@@

Definition linv : D → option C :=
  prec C 
    ...
    (fun (aᵢ : Aᵢ) (bᵢ' : Bᵢ aᵢ → option C) ⇒
       match (detectable bᵢ') with
       | left h ⇒
           let (bᵢ, _) := unlift bᵢ' h in Some (cᵢ aᵢ bᵢ)
       | right _ ⇒ None
       end)
    ... .
\end{minted}
Since each ``constructor'' in "C" and "D" are well-correpsonded, when invoking "C_rect" and "prec", each case is handled similarly, and they only vary along the $\{ "A"_i \vdash "B"_i \}_{i}$ types.

The definition of "inj" and "linv" is well-formed but their behaviour is still burried in the definition. The following two equations can clarify them.

@@

Finally,  these two equations is enough to prove that "linv" is really a left inverse, via induction
\begin{minted}{Coq}
linv (inj (cᵢ a b)) 
= linv (dᵢ a (fun x ⇒ inj (b x)))         (* by inj_compute *)
= Some (cᵢ a (fun x ⇒ linv (inj (b x))))  (* by linv_compute *)
= Some (cᵢ a (fun x ⇒ b x))              (* by I.H. *)
= Some (cᵢ a b)                           (* by eta rule *)
\end{minted}
\end{proof}

@@

\ref{prop:detectable-partiality} is not generally supported when using function (infinitary branching) in the constructors, for example, the case of using \textit{higher order abstract syntax} to represent functions in "tm". However, in those cases, we still have partial recursor provable and useful---we just cannot see the partial recursor as a strong toolkit anymore.

We emphasize that in \cref{thm:prec-complete}, each function $"d"_i$ can
actually be considered as a constructor, because they can be reflected
to real constructors of "C" using the left inverse "linv". Thus, \cref{thm:prec-complete}
implies that (1) every future extension of the inductive type can
support this partial recursion (trivially); (2) every type supporting
this partial recursor with its computational axioms at least supports
these constructors because of the embedding. In other words, \textit{a
type (at least) supports these constructors if and only if this type
supports the corresponding partial recursor (with the computational
axioms)}.  Thus, we can argue that the partial recursor gives a sound and
complete extensional characterization of all the extensions of a given (non-indexed)
inductive type.\YZ{What about indexed inductive types?}\EDJreply{Done. I also have a formalization of this proof now, for indexed type and dependent eliminator. Though the formalization is about single constructor.}

We can support partial dependent eliminator as well, but partial dependent eliminator can deduce "prec" and our "prec" is enough to contruct this left inverse for (non-indexed) inductive type. 
\begin{minted}{Coq}
  pdelim : ∀ (P : D → Set), .., 
  (rᵢ : (∀ (x : Aᵢ) (w : Bᵢ x → D), (∀ (b : Bᵢ x), option (P (w b))) 
    → option (P (dᵢ x w)))),  .., ∀ (d : D) → option (P d)
\end{minted}
To generalize to indexed inductive type, we need to base on indexed W type~\cite{martin1982constructive, morris2009indexed,jashug2017}. We have mechanized the argument for indexed W type and dependent eliminator, but for the case with only one constructor in "C" and "D". This mechanization\footnote{\url{https://gist.github.com/DKXXXL/9b4e48350b966fa3739cfb71f1a9a852}} is a direct generalization of the proof for \cref{thm:prec-complete}.

As mentioned, "tm_prect_STLC" is axiomatized along with
four equalities describing its computational behaviors, one for each constructor.
After unfold the definition of "tmPR", the equality for constructor "tm_abs" is as follows:

>>>>

As analyzed earlier, ordinary recursors, such as "tm_rect", are impossible
within a family in which the name of the inductive type is late bound.
However,
a key observation is that extensible inductive types still admit a
weakened elimination principle where the motive is an \lsti{option} type.
For example, within family "STLC", the partial recursor for
"tm" has the following type, with extra equations describing
its behavior on each constuctor:

\begin{centered}
\begin{minipage}{\textwidth}
\newcommand\hlcase[1]{\tikzmarkin[disable rounded corners=true,set fill color=yellow!18,set border color=yellow!18]{#1}(0.00,-0.05)(-0.00,0.20)}
\begin{lstlisting}[basicstyle=\fontsize{8.25}{9.9}\ttfamily]
tm_prect_STLC : \forall (P : tm -> Type), [@\hlcase{tm-partialrec-case-unit}@]option (P tm_unit)[@\tikzmarkend{tm-partialrec-case-unit}@] -> [@\hlcase{tm-partialrec-case-var}@](\forall x, option (P (tm_var x)))[@\tikzmarkend{tm-partialrec-case-var}@] ->
  [@\hlcase{tm-partialrec-case-abs}@](\forall x t, option (P t) -> option (P (tm_abs x t)))[@\tikzmarkend{tm-partialrec-case-abs}@] ->
  [@\hlcase{tm-partialrec-case-app}@](\forall t1, option (P t1) -> \forall t2, option (P t2) -> option (P (tm_app t1 t2)))[@\tikzmarkend{tm-partialrec-case-app}@] -> \forall t, option (P t)
(* For arbitrary [@$P$@] [@$r_1$@] [@$r_2$@] [@$r_3$@] [@$r_4$@], and we denote [@"tmPR"@] := [@"tm_prect_STLC"@] [@$P$@] [@$r_1$@] [@$r_2$@] [@$r_3$@] [@$r_4$@] *)
(* We will have the following equations *)
tm_unit_eq_STLC : tmPR tm_unit      = [@$r_1$@]
tm_var_eq_STLC  : tmPR (tm_var x)   = [@$r_2$@] x
tm_abs_eq_STLC  : tmPR (tm_abs x b) = [@$r_3$@] a b (tmPR b)
tm_app_eq_STLC  : tmPR (tm_app a b) = [@$r_4$@] a (tmPR a) b (tmPR b)
\end{lstlisting}
\end{minipage}
\end{centered}

"tm_prect_STLC" cannot be a trivial function even though its return type has "option". Because these four equations uniquely describe the "tm_prect_STLC" on the four constructors and make "tm_prect_STLC" behave similarly to the ordinary dependent eliminator of the non-extensible version of "tm".

@@

This partial recursor seems non-canonical compared to the (dependent) eliminator for inductive types as the latter can be considered a universal property. This extensionality makes dependent eliminator one of \textit{the most powerful} reasoning tools around inductive type.

Even though we cannot prove partial recursors are \textit{the
most powerful} reasoning tool for extensible inductive types,
the fact that propositional partial
recursors can derive injectivity already demonstrate itself as a good \textit{extensional characterization} of
(non-indexed) extensible inductive type (under certain
constraints)---because we can ``embed'' vanilla inductive types into
types supporting a partial recursor: a bit more formally speaking, and
restricting our focus to non-indexed inductive types:

@@

  \begin{itemize}
    \item \textlabel{(Detectable Partiality)}{prop:detectable-partiality} for each pair $x : A_i \vdash B_i(x)$,
    \begin{minted}{Coq}
      Axiom detectable:
      forall {T} {a} (f: Bᵢ a → option T),
        {forall x, f x <> None} + {exists x, f x = None}.
    \end{minted} 
    \item Define $"C"$ as the inductive type using this list, i.e.,
    \begin{minted}{Coq}
Inductive C : Set := ... | cᵢ : forall (x : Aᵢ), (Bᵢ x → C) → C | ...
    \end{minted}
    \item Assume an arbitrary type \mintinline{Coq}{D : Set} with
    \begin{itemize}
      \item $n$ \textbf{functions} \mintinline{Coq}{dᵢ : forall (x : Aᵢ), (Bᵢ x → D) → D}  
      \item a partial recursor
      \begin{minted}[escapeinside=@@]{Coq}
prec : forall (R : Set), ..,
  (rᵢ : (forall (x : Aᵢ), (Bᵢ x → option R) → option R)), ..,@\YZ{Can R be indexed?}\EDJreply{This part actually currently incorrect and needs an extra constraint. Let me fix other parts first}\YZreply{Is this constraint Detectable Partiality?}@
      D → option R
      \end{minted}
      \item and $n$ (propositional) computational axioms: for all $i$, 
      \begin{minted}{Coq}
prec {R} r₁ r₂ .. rₙ (dᵢ aᵢ bᵢ) 
    = rᵢ aᵢ (fun x ⇒ prec {R} r₁ r₂ .. rₙ (bᵢ x))
      \end{minted}
    % where \mintinline{Coq}{(lift rᵢ) : (forall (x : Aᵢ), (Bᵢ x -> option R) -> option R)} is defined as expected
    \end{itemize}
  \end{itemize}
  Then there exists an \textbf{embedding} from "C" to "D".
  More concretely, we can have \mintinline{Coq}{inj : C → D} and 
  \mintinline{Coq}{linv : D → option C} defined using the eliminator of\/
  "C" and "prec" of\/ "D" (both acting like identity) such that
  \begin{minted}[escapeinside=@@]{Coq}
    forall c : C, linv (inj c) = some c@\YZ{Does 'linv' correspond to 'reflect' above? If so, then the reader should be explicitly reminded of this fact rather than having to guess it.}\EDJreply{Added below. I mentioned left inverse 'linv' is the generalization}@
  \end{minted}
\end{theorem}
\begin{proof} 
First, for a fixed \mintinline{Coq}{x : Aᵢ}, given \mintinline{Coq}{fᵢ :  Bᵢ x → T},
\\ we have \mintinline{Coq}{lift fᵢ : Bᵢ x → option T}, simply wrapping option around the returning result.

@@

Definition linv : D → option C :=
  prec C 
    ...
    (fun (aᵢ : Aᵢ) (bᵢ' : Bᵢ aᵢ → option C) ⇒
       match (detectable bᵢ') with
       | left h ⇒
           let (bᵢ, _) := unlift bᵢ' h in Some (cᵢ aᵢ bᵢ)
       | right _ ⇒ None
       end)
    ... .
\end{minted}
Since each ``constructor'' in "C" and "D" are well-corresponded, when invoking "C_rect" and "prec", each case is handled similarly, and they only vary along the $\{ "A"_i \vdash "B"_i \}_{i}$ types.

The definition of "inj" and "linv" is well-formed but their behavior is still buried in this complicated formulation. The following two equations can clarify them.

@@

Finally,  these two equations are enough to prove that "linv" is really a left inverse, via induction
\begin{minted}{Coq}
linv (inj (cᵢ a b)) 
= linv (dᵢ a (fun x ⇒ inj (b x)))         (* by inj_compute *)
= Some (cᵢ a (fun x ⇒ linv (inj (b x))))  (* by linv_compute *)
= Some (cᵢ a (fun x ⇒ b x))              (* by I.H. *)
= Some (cᵢ a b)                           (* by eta rule *)
\end{minted}
\end{proof}

@@

\ref{prop:detectable-partiality} is not generally supported when using a function (infinitary branching) in the constructors, for example, the case of using \textit{higher order abstract syntax} to represent functions in "tm". However, in those cases, we still have the partial recursor provable and useful---we just cannot see the partial recursor as a strong toolkit anymore.

We emphasize that in \cref{thm:prec-complete}, each function $"d"_i$ can
actually be considered a constructor because they can be reflected
to real constructors of "C" using the left inverse "linv". Thus, \cref{thm:prec-complete}
implies that (1) every future extension of the inductive type can
support this partial recursion (trivially); (2) every type supporting
this partial recursor with its computational axioms at least supports
these constructors because of the embedding. In other words, \textit{a
type (at least) supports these constructors if and only if this type
supports the corresponding partial recursor (with the computational
axioms)}.  Thus, we can argue that the partial recursor gives a sound and
complete extensional characterization of all the extensions of a given (non-indexed)
inductive type.\YZ{What about indexed inductive types?}\EDJreply{Done. I also have a formalization of this proof now, for indexed type and dependent eliminator. Though the formalization is about single constructor.}

We can support partial dependent eliminator as well, but partial dependent eliminator can deduce "prec" and our "prec" is enough to construct this left inverse for (non-indexed) inductive type. 
\begin{minted}{Coq}
  pdelim : ∀ (P : D → Set), .., 
  (rᵢ : (∀ (x : Aᵢ) (w : Bᵢ x → D), (∀ (b : Bᵢ x), option (P (w b))) 
    → option (P (dᵢ x w)))),  .., ∀ (d : D) → option (P d)
\end{minted}
To generalize to indexed inductive type, we need to base on indexed W type~\cite{martin1982constructive, morris2009indexed,jashug2017}. We have mechanized the argument for indexed W type and dependent eliminator, but for the case with only one constructor in "C" and "D". This mechanization\footnote{\url{https://gist.github.com/DKXXXL/9b4e48350b966fa3739cfb71f1a9a852}} is a direct generalization of the proof for \cref{thm:prec-complete}.

As mentioned earlier, "tm_prect_STLC" is axiomatized along with
four equalities describing its computational behaviors, one for each constructor.
After unfolding the definition of "tmPR", the equality for constructor "tm_abs" is as follows:
